---
title: "Case Study"
author: "Ojas"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = list(CRAN="http://cran.rstudio.com/"))
```
## Installing and loading common packages and libraries
we will load all the necessary libraries required to process, analyze, and visualize the data.

```{r echo=TRUE}
install.packages('tidyverse')
install.packages('dplyr')
install.packages('janitor')
install.packages('skimr')
library(dplyr)
library(janitor)
library(skimr)
library(tidyverse)
```


## Loading CSV files
We will use *dailyActivity_merged.csv*, *sleepDay_merged.csv*, and *dailySteps_merged.csv* datasets in this analysis. 

```{r}
daily_activity <- read.csv("dailyActivity_merged.csv")
sleep_day <- read.csv("sleepDay_merged.csv")
daily_steps <- read.csv("dailySteps_merged.csv")
```

## Exploring a few key tables
We will now look into our datasets
*daily_activity*
```{r}
head(daily_activity)
```

```{r}
glimpse(daily_activity)
```

```{r}
colnames(daily_activity)
```

*sleep_day*
```{r}
head(sleep_day)
```

```{r}
glimpse(sleep_day)
```

```{r}
colnames(sleep_day)
```
*daily_steps*
```{r}
head(daily_steps)
```

```{r}
glimpse(daily_steps)
```

```{r}
colnames(daily_steps)
```
Note that all datasets have the 'Id' field - this can be used to merge the datasets.

If you look into the daily_activity dataset you will see data of daily_steps is already present in daily_activity$TotalSteps. This is the reason why other datasets were not loaded.

## Understanding some summary statistics

We will use *n_distinct()* and *nrow* to count unique values and number of rows respectively. As Id is the only column that can used to identify different users, we will use it to calculate number of unique users.

```{r distinct users}
n_distinct(daily_activity$Id)
n_distinct(sleep_day$Id)
```
This shows that *daily)activity* contains information of 33 users whereas *sleep_day* contains 24 unique users.
As for the number of observations in each datasets
```{r observations}
nrow(daily_activity)
nrow(sleep_day)
```
*summary()* is used to to summarize and give your key insights from the data.

Here in the *daily_activity* dataset:
```{r}
daily_activity %>%  
  select(TotalSteps,
         TotalDistance,
         SedentaryMinutes,VeryActiveMinutes,Calories) %>%
  summary()
```
The summary of *daily_activity* shows that the average steps taken in a day by a user is 7406, which is more than the minimum recommended steps per day but less than recommended steps per day for health by CDC. Looking the SedentaryMinutes we get to know that an average person is sitting idle and not doing any physical activity for about 16.52 hours(991.2 minutes), which is very high. On average CDC recommends a person to do vigorous activity for atleast 75 minutes a week, from the dataset we can see that on average 21.16 minutes is the time spent to do vigorous activity which acounts to 148.12 minutes. So we can say on average a fitbit user is doing more activity than recommended and availing more health benefits. Furthermore, on average a fitbit user is burning 2304 calories per day which depends is good or not on your goal. To lose weight on average 3500 calories needs to be burned according to CDC. On average a person burns 1800 calories a day, in that sense a fitbit user is doing better.

For the *sleep_day* dataset:
```{r}
sleep_day %>%  
  select(TotalSleepRecords,
  TotalMinutesAsleep,
  TotalTimeInBed) %>%
  summary()
```

The summary of *sleep_day* shows that on an average a fitbit user is sleeping for 419.5 minutes(6.99 hours) which is just touches the recommended sleep of 7 hours in a day. If you look at the time spent on bed, average time comes out to be 458.6 which means that a person is spending 39.1 minutes laying on bed. According to Health Central, people should not spend more than 1 hour in bed awake.

## Merging these two datasets together

We will use Inner Join to combine these two datasets to gain more insight on the relation between daily activity and sleep cycle of these fibit users
```{r}
combined_data <- merge(sleep_day, daily_activity, by="Id")
```

Take a look at how many participants are in this data set.

```{r}
n_distinct(combined_data$Id)
```
As we used Inner join, the users whose sleep cycle information were not recorded are excluded from the combined dataset. We can use Outer Join if we want to include those data in the combined dataset.
```{r}
glimpse(combined_data)
```

```{r}
combined_data %>% 
  select(TotalSteps,SedentaryMinutes,VeryActiveMinutes,TotalMinutesAsleep,Calories) %>% 
  summary()
```
Looking at *combined_data* we get to know that on average more the active minutes spent by an user more is their minutes asleep. That is the more you burn calories better gets your sleep cycle.

## Visualizing to gain insights

We will use *ggplot()* from ggplot2 to visualize our data and look into the pattern and trends in the dataset.

```{r}
ggplot(data=daily_activity, aes(x=TotalSteps, y=Calories)) + geom_point() +
  stat_smooth(method = lm) + 
  labs(title = "Relation between Total Steps and Calories burnt in a day")
```
As we can see there is a positive relation between Total steps and Calories burnt, which tells us that more steps taken by a fitbit user, more calories were burnt by them.

```{r}
ggplot(data=daily_activity, aes(x=VeryActiveMinutes,y=Calories)) + geom_point() +
  stat_smooth(method = lm) + 
  labs(title = "Relation between Very Active Minutes and Calories burnt")
```
The relation between Active minutes and Calories burn tis positive meaning more a person spends doing activity more calories will be burnt by that person.

```{r}
ggplot(data = daily_activity, aes(x=SedentaryMinutes,y=Calories)) + geom_point() +
  stat_smooth(method = lm) + 
  labs(title = "Relation between Sedentary Minutes and Calories burnt")
```
The relation between Sedentary minutes and Calories burnt comes to be almost linear inclined toward negetive side. This means that if a user is idle the calories burnt will not be affected much but as the idle time increase the calories burnt will be less than that of a person who is not idle for that amount of time.

```{r}
ggplot(data =sleep_day, aes(x=TotalMinutesAsleep,y=TotalTimeInBed)) + geom_point() +
  stat_smooth(method = lm) + 
  labs(title = "Relation between Total minutes asleep and Total time in bed")
```
This visualization tells uss the relation etween Total time asleep and Total time in bed. As most of the points are near the line, it tells us that most of the time time spent on bed is similar to time asleep.